# Reproducible Research: Peer Assessment 1

We are going to analyse data collected from personal activity monitoring device, answerings a serie of questions.

Data is stored in a dataset that includes this variables:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

## Loading and preprocessing the data

We can read the data set from a comma-separated-value (CSV) file, zipped in "activity.zip" file. 

Steps to get data:

1. unzip data
2. read data from CVS file
3. Process data to replace NA values with 0. Note: we'll need original data later


``` {r echo = TRUE}
## loadl library
library(lattice)

## unzip data
unzip("activity.zip")

## read data
data <- read.csv("activity.csv")

## replace NA values with 0 
## Note: we'll need original data later
datafiltered <- na.omit(data)

```

## What is mean total number of steps taken per day?

We'll make a histogram of the total number of steps taken each day:

``` {r echo = TRUE}

## sum number of steps, grouping by days
stepsbyday <- aggregate(datafiltered$steps, by=list(datafiltered$date), sum)

## change names of columns in dataframe
names(stepsbyday) <- c("day","number_of_steps")

hist(stepsbyday$number_of_steps, xlab = "Number of steps taken each day",
     main="Number of steps taken each day")

```

The mean and median total number of steps taken per day are:

``` {r echo=TRUE}
cat("Mean number of steps per day: ",mean(stepsbyday$number_of_steps))
cat("Median number of steps per day: ",median(stepsbyday$number_of_steps))
```

## What is the average daily activity pattern?

To gain insight into this question, we'll make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis):

``` {r echo = TRUE}
## calculate mean number of steps by interval
stepsinterval <- aggregate(datafiltered$steps, 
                           by=list(datafiltered$interval), mean)

## change names of columns in dataframe
names(stepsinterval) <- c("interval","number_of_steps")

## plot number of steps by interval 
plot(stepsinterval$interval, stepsinterval$number_of_steps, type="l", 
     xlab="Interval", ylab="Number of steps")
```

The daily pattern shows a sharp peak of activity. Then activity quikly declines, and grow more slowly in remainder intervals. 

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
maxinterval<- stepsinterval[max(stepsinterval$number_of_steps) == 
                                    stepsinterval$number_of_steps,1]
cat("Interval with maximun number of steps: ", maxinterval)
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset:

``` {r echo=TRUE}
cat("Total number of missing values: ", colSums(is.na(data))[1])
```

Devise a strategy for filling in all of the missing values in the dataset:

- Our strategy: We'll create a new dataset named "dataf", <b> replacing missing values with mean values for each interval</b> :

```{r echo=TRUE}
## merge with data frame containing mean number of steps 
dataf <- merge(data,stepsinterval)
## replace NAs with median number of steps per interval
dataf$steps[is.na(dataf$steps)] <- dataf$number_of_steps[is.na(dataf$steps)]
```

Then, we make a histogram of the total number of steps taken each day, using "dataf" dataframe:

```{r echo=TRUE}
## sum number of steps, grouping by days
newstepsbyday <- aggregate(dataf$steps, by=list(dataf$date), sum)

## change names of columns in dataframe
names(newstepsbyday) <- c("day","number_of_steps")

## plot histogram
hist(newstepsbyday$number_of_steps, xlab = "Number of steps taken each day",
     main="Number of steps taken each day after missing values were imputed")
```

And finally we calculate mean and median:

``` {r echo=TRUE}
cat("Mean number of steps per day: ",mean(newstepsbyday$number_of_steps))
cat("Median number of steps per day: ",median(newstepsbyday$number_of_steps))
```

The results match original (non-corrected) data.

Imputing missing data doesn't change overall conclussions.

## Are there differences in activity patterns between weekdays and weekends?

This plot will help us to respond the question:

```{r echo=TRUE}
## convert to date
dataf$date <- as.Date(dataf$date)

## add new column: weekend or weekday? 
dataf$day <- rep("weekday",nrow(dataf))
dataf$day[ as.POSIXlt(dataf$date)$wday %in% c(6,7) ] <- "weekend"

## sum steps by interval and day of week
dataf <- aggregate(dataf$steps, by=list(dataf$interval, dataf$day), mean)
names(dataf) <- c("interval","day", "mean_number_steps")

## plot graph
xyplot(mean_number_steps ~ interval | day, data = dataf, layout = c(1, 2),type="l")
```

Weekdays and weekend days show different patterns.

Weekdays, activity tends to concentrate in the first half of day (intervals from 500 to 1000).  

But activity on weekend shows a more irregular pattern, with relevant peaks on the afternoon.
